

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # loads the tidyverse tools
library(DBI) # loads our database interface
library(RPostgres) # loads the database driver for PostgreSQL
library(connections) # helps RPostgres work with RStudio

con <- connection_open(RPostgres::Postgres(),
          dbname = "aact",
          host = "aact-db.ctti-clinicaltrials.org",
          user = "username",
          password = "password",
          # Tell the driver to return very large integers as floating point (vs truncating them)
          bigint = "numeric")
```
## Clinical Trial Attrition in Neuorpsychiatric Clinical Trials Using Machine Intelligence

### Introduction
Clinical trial participation retention directly affects the generalizability and validity of the results of that trial. The primary goal of this notebook is to use a data driven approach to investigate the predictive features leading to patient attrition in Alzheimer's Disease clinical trials. The data source is aggregated at the trial level. 

#### Alzheimer's Disease
Create list of nct_ids that have some form of "alzheimer" listed as downcase_name of the study from the conditions table. Found 2477 studies that matched the syntax below.  

##### Method for getting Alzheimer's Disease studies :  
  - Table : ctgov.conditions   
  - Column : downcase_name  
    - Strings Pulled : 'alzheimers'  
    
###### Alzheimer's Disease study key method
```{sql connection=con}
DROP TABLE IF EXISTS alzheimers_study_key_V2;
```

```{sql connection=con}
WITH alzheimers_studies AS (
  SELECT *
  FROM ctgov.conditions 
  WHERE downcase_name LIKE '%alzheimer%' AND downcase_name NOT LIKE 'non-alzheimer degenerative dementia'
  ), enrollment_info AS ( 
    SELECT studies.nct_id, SUM(enrollment) AS enrollment_total
    FROM ctgov.studies
    GROUP BY nct_id
 ), dropout_info AS (
    SELECT drop_withdrawals.nct_id, SUM(count) AS drop_count_total
    FROM ctgov.drop_withdrawals
    GROUP BY nct_id
 )
SELECT alzheimers_studies.nct_id, enrollment_total, drop_count_total
INTO TEMP alzheimers_study_key_V2
FROM alzheimers_studies 
LEFT JOIN enrollment_info
ON alzheimers_studies.nct_id = enrollment_info.nct_id
LEFT JOIN dropout_info
ON alzheimers_studies.nct_id = dropout_info.nct_id
```

```{sql connection=con}
SELECT COUNT(DISTINCT nct_id)
FROM alzheimers_study_key_V2;
```    


## create table with ae nct_ids, subject_count, and ids for creating adverse events table
```{sql connection=con, output.var="alzheimers_ae_key"}
WITH alzheimers_studies AS (
  SELECT *
  FROM ctgov.conditions 
  WHERE downcase_name LIKE '%alzheimer%' AND downcase_name NOT LIKE 'non-alzheimer degenerative dementia'
  ), reported_events AS (
    SELECT reported_events.nct_id, reported_events.subjects_affected, reported_events.id
    FROM ctgov.reported_events
 )
SELECT alzheimers_studies.nct_id, reported_events.id, subjects_affected
FROM alzheimers_studies 
LEFT JOIN reported_events
ON alzheimers_studies.nct_id = reported_events.nct_id;
```

```{r}
write.csv(alzheimers_ae_key, "/Users/nataliecortopassi/Library/Mobile Documents/com~apple~CloudDocs/510 - HIDS/Adverse Events/ae_key_dfs/alzheimers_ae_key.csv")
```    

##### Create Seperate events table with counts per event type
  - Table : ctgov.reported_event_totals
  - Columns 
      - event_type *should be four "death, serious, other, or nan" for each study with counts for each*
          - Indicates the type of event (options are "Serious", "Other" and "Deaths")
      - subjects_affected (count_event_subjects) *calculated by adding all values grouped by nct_id and event_type*
          - Number of subjects affected by the event
```{sql connection=con}
DROP TABLE IF EXISTS alzheimers_event_table;
```

```{sql connection=con}
WITH event_info AS (
    SELECT reported_event_totals.nct_id, reported_event_totals.event_type, SUM(subjects_affected) AS subjects_affected_total
    FROM ctgov.reported_event_totals
    GROUP BY nct_id, event_type
 ), alzheimers_key AS ( 
    SELECT alzheimers_study_key_V2.nct_ID
    FROM alzheimers_study_key_V2
)
 SELECT alzheimers_key.nct_id, subjects_affected_total, event_type
 INTO TEMP alzheimers_event_table
 FROM alzheimers_key 
 LEFT JOIN event_info
 ON alzheimers_key.nct_id = event_info.nct_id
``` 

```{sql connection=con, output.var="alzheimers_event_table"}
SELECT * 
FROM alzheimers_event_table;
```

```{r}
write.csv(alzheimers_event_table, "/Users/nataliecortopassi/Library/Mobile Documents/com~apple~CloudDocs/510 - HIDS/alzheimers_df/alzheimers_event_table.csv")
```
    
#### Select Features and Columns to create target outcome (percentage dropped) - V2
Create table that matches the nct_id found above (alzheimers studies) and add features from other aact tables

##### Feature Selection
  - Table : ctgov.drop_withdrawals
  - Columns 
      - count (number_dropped) *calculated by adding all values grouped by nct_id*
          - Number of participants (and units, if applicable) that did not complete the study or period. This is calculated automatically by subtracting Completed from started. 
  - Table : ctgov.studies
  - Columns 
      - enrollment (number_enrolled) *calculated by adding all values grouped by nct_id*
          - The estimated total number of participants to be enrolled (target number) or the actual total number of participants that are enrolled in the clinical study.
      - study_type
          - The nature of the investigation or investigational use for which clinical study information is being submitted. Select one.
              - Interventional (clinical trial): Participants are assigned prospectively to an intervention or interventions according to a protocol to evaluate the effect of the intervention(s) on biomedical or other health related outcomes.
              - Observational: Studies in human beings in which biomedical and/or health outcomes are assessed in pre-defined groups of individuals. Participants in the study may receive diagnostic, therapeutic, or other interventions, but the investigator does not assign specific interventions to the study participants. 
                  - Patient Registry: An observational study that is also considered to be a Patient Registry. This type of study should only be registered once in the Protocol Registration and Results System (PRS), by the sponsor responsible for the primary data collection and analysis.
              - Expanded Access: An investigational drug product (including biological product) available through expanded access for patients who do not qualify for enrollment in a clinical trial. Expanded Access includes all expanded access types under section 561 of the Federal Food, Drug, and Cosmetic Act: (1) for individual patients, including emergency use; (2) for intermediate-size patient populations; and (3) under a treatment IND or treatment protocol. 
      - phase (study_phase)
          - Early Phase
          - Phase 1 
          - Phase 1/2
          - Phase 2
          - Phase 2/3
          - Phase 3
          - Phase 4
          - Not Applicable 
          - nan
      - number_of_arms 
          - The number of arms in the clinical trial. For a trial with multiple periods or phases that have different numbers of arms, the maximum number of arms during all periods or phases.
      - has_expanded_access
          - True (has expanded access) / False (does not have expanded access)
      - has_dmc
          - Indicate whether a data monitoring committee has been appointed for this study. The data monitoring committee (board) is a group of independent scientists who are appointed to monitor the safety and scientific integrity of a human research intervention, and to make recommendations to the sponsor regarding the stopping of the trial for efficacy, for harms or for futility.
          - True (has dmc) / False (does not have dmc)
  - Table : ctgov.calculated_values
  - Columns 
      - number_of_facilities
          - Number of participating facilities in a clinical study
      - actual_duration (study_duration_months)
          - Number of months between start_date & primary_completion_date
      - number_of_sae_subjects
          - Count number of related serious event rows
      - minimum_age_num
          - Convert minimum age to an integer
              - transform this into binary - "prediatric" / "non-pediatric"
  - Table : ctgov.designs
  - Columns 
      - intervention_model (intervention_model_type)
          - The strategy for assigning interventions to participants.
              - Single Group: Clinical trials with a single arm
              - Parallel: Participants are assigned to one of two or more groups in parallel for the duration of the study
              - Crossover: Participants receive one of two (or more) alternative interventions during the initial phase of the study and receive the other intervention during the second phase of the study
              - Factorial: Two or more interventions, each alone and in combination, are evaluated in parallel against a control group
              - Sequential: Groups of participants are assigned to receive interventions based on prior milestones being reached in the study, such as in some dose escalation and adaptive design studies
      - allocation (allocation_type)
          - The method by which participants are assigned to arms in a clinical trial.
              - N/A (not applicable): For a single-arm trial
              - Randomized: Participants are assigned to intervention groups by chance
              - Nonrandomized: Participants are expressly assigned to intervention groups through a non-random method, such as physician choice
      - masking (masking_type)
          - The party or parties involved in the clinical trial who are prevented from having knowledge of the interventions assigned to individual participants.
              - Roles, if Masking:
                  - Participant
                  - Care Provider
                  - Investigator
                  - Outcomes Assessor: The individual who evaluates the outcome(s) of interest
              - No Masking
  - Table : ctgov.eligibilities
  - Columns 
      - gender (study_gender_eligibility)
          - All, Female, Male, and nan as options for gender eligibility
      - healthy_volunteers (health_status_eligibility)
          -  Indication that participants who do not have a disease or condition, or related conditions or symptoms, under study in the clinical study are permitted to participate in the clinical study. 
              - accepts healthy volunteers- "yes" / "no"
  - Table : ctgov.interventions
  - Columns 
      - intervention_type
          - Drug: Including placebo
          - Device: Including sham
          - Biological/Vaccine
          - Procedure/Surgery
          - Radiation
          - Behavioral: For example, psychotherapy, lifestyle counseling
          - Genetic: Including gene transfer, stem cell and recombinant DNA
          - Dietary Supplement: For example, vitamins, minerals
          - Combination Product: Combining a drug and device, a biological product and device; a drug and biological product; or a drug, biological product, and device
          - Diagnostic Test: For example, imaging, in-vitro
          - Other
      - name (intervention_name)
          - A brief descriptive name used to refer to the intervention(s) studied in each arm of the clinical study. A non-proprietary name of the intervention must be used, if available. If a non-proprietary name is not available, a brief descriptive name or identifier must be used.


```{sql connection=con}
DROP TABLE IF EXISTS V2_alzheimers;
```

```{sql connection=con}
WITH alzheimers_list_key AS (
  SELECT *
  FROM alzheimers_study_key_V2 
), enrolled_count_list AS (
  SELECT * 
  FROM ctgov.baseline_counts 
), dropped_count_list AS (
  SELECT * 
  FROM ctgov.drop_withdrawals
), study_info AS (
  SELECT * 
  FROM ctgov.studies
), calculated_info AS (
  SELECT * 
  FROM ctgov.calculated_values
), country_info AS (
  SELECT * 
  FROM ctgov.countries
), design_info AS (
  SELECT * 
  FROM ctgov.designs
), eligibility_info AS (
  SELECT * 
  FROM ctgov.eligibilities
), intervention_info AS (
  SELECT * 
  FROM ctgov.interventions
), reported_events_info AS (
  SELECT * 
  FROM ctgov.reported_event_totals
)

SELECT alzheimers_list_key.nct_id, alzheimers_list_key.enrollment_total, alzheimers_list_key.drop_count_total, study_info.study_type, study_info.phase AS study_phase, study_info.number_of_arms, study_info.has_expanded_access, study_info.has_dmc, calculated_info.number_of_facilities,  calculated_info.actual_duration AS study_duration_months, calculated_info.number_of_sae_subjects, calculated_info.minimum_age_num, design_info.intervention_model AS intervention_model_type, design_info.allocation AS allocation_type, design_info.masking AS masking_type, eligibility_info.gender AS study_gender_eligibility, eligibility_info.healthy_volunteers AS health_status_eligibility
    INTO TEMP V2_alzheimers
    FROM alzheimers_list_key
    LEFT JOIN enrolled_count_list
    ON alzheimers_list_key.nct_id = enrolled_count_list.nct_id
    LEFT JOIN dropped_count_list
    ON alzheimers_list_key.nct_id = dropped_count_list.nct_id
    LEFT JOIN study_info
    ON alzheimers_list_key.nct_id = study_info.nct_id
    LEFT JOIN calculated_info
    ON alzheimers_list_key.nct_id = calculated_info.nct_id
    LEFT JOIN country_info
    ON alzheimers_list_key.nct_id = country_info.nct_id
    LEFT JOIN design_info
    ON alzheimers_list_key.nct_id = design_info.nct_id
    LEFT JOIN eligibility_info
    ON alzheimers_list_key.nct_id = eligibility_info.nct_id
    LEFT JOIN intervention_info
    ON alzheimers_list_key.nct_id = intervention_info.nct_id
    LEFT JOIN reported_events_info
    ON alzheimers_list_key.nct_id = reported_events_info.nct_id
    GROUP BY alzheimers_list_key.nct_id, enrollment_total, drop_count_total, study_type, study_phase, number_of_arms, has_expanded_access, has_dmc, number_of_facilities, study_duration_months, number_of_sae_subjects, minimum_age_num, intervention_model_type, allocation_type, masking_type, study_gender_eligibility, health_status_eligibility;
```

```{sql connection=con, output.var="V2_alzheimers"}
SELECT * 
FROM V2_alzheimers;
```

```{r}
write.csv(V2_alzheimers, "/Users/nataliecortopassi/Library/Mobile Documents/com~apple~CloudDocs/510 - HIDS/alzheimers_df/V2_alzheimers.csv")
```   

#### Create treatment_type table
```{sql connection=con, output.var="alzheimers_treatment_type"}
SELECT V2_alzheimers.nct_id, interventions.intervention_type
FROM V2_alzheimers
LEFT JOIN ctgov.interventions
ON V2_alzheimers.nct_id = interventions.nct_id
```

```{sql connection=con}
SELECT DISTINCT(intervention_type)
FROM V2_alzheimers
```

```{r}
write.csv(alzheimers_treatment_type, "/Users/nataliecortopassi/Library/Mobile Documents/com~apple~CloudDocs/510 - HIDS/alzheimers_df/alzheimers_treatment_type.csv")
```    
    
  
#### Select dropout reasons
Create table of dropout reasons with nct_id associated with alzheimers studies
```{sql connection=con, output.var="alzheimers_dropout_reason"}
SELECT drop_withdrawals.reason AS dropout_reason, V2_alzheimers.nct_id
FROM V2_alzheimers
LEFT JOIN ctgov.drop_withdrawals
ON V2_alzheimers.nct_id = drop_withdrawals.nct_id;
```
    
```{r}
write.csv(alzheimers_dropout_reason, "/Users/nataliecortopassi/Library/Mobile Documents/com~apple~CloudDocs/510 - HIDS/alzheimers_df/alzheimers_dropout_reason.csv")
```    
        
    
    
    
    
